### 2018-03-02 manuscript.rmd
changes throughout text & to bib file

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at all depths of the ocean - from the surface, like the familiar Portuguese man of war, to the deep sea. Siphonophores play important roles in ocean ecosystems, and are among the most abundant gelatinous predators. A previous phylogenetic study based on two ribosomal RNA genes provided insight into the internal relationships between major siphonophore groups, however there was little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny based on new transcriptome data from 29 siphonophore species analyzed in combination with 14 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including sexual system (monoecy vs. dioecy), gain and loss of zooid types, life history traits, and habitat. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but we find strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose at least twice.

@@

Siphonophores (Fig. 1 and 2) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical; @purcell1981dietary; @pages2001gelatinous;  @Choy20172116]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at all depths in the ocean. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be more than 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, Rhodaliidae [@pugh1983benthic], that are tethered to the bottom for part of their lives. There are currently 187 valid described siphonophore species.

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years due to their unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones (Fig. 2) then arise and asexually produce other genetically identical zooids that remain attached [@carre1969etude; @carre1991complete; @carre1967; @carre1995ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from other colonial animals in their colony structure and development -- their zooids are highly functionally specialized and arranged in precise, repeating, species-specific patterns [@Beklemishev1969;@cartwright2010character]. Zooids are specialized for a range of functions, including feeding, reproduction, or swimming (Fig. 2) [@dunn2006evolution].

@@

The phylogenetic relationships recovered in this study received strong support across analysis methods (Fig. 3A), with a couple of localized exceptions (Fig. 3B). Most clades are consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

@@

## 3.2 Character Evolution
```{r char_evo_setup, include=F, echo=FALSE}
#Load data
read.csv('character_coding/main_characters.csv', header = T, sep = ',') -> cdata
rownames(cdata) = cdata$Species

@@

### 3.3 Evolution of Monoecy

@@

Ancestral character state reconstructions favor the hypothesis that monoecy arose twice (Fig. 5A), once in Calycophorae and once in Clade A. This is consistent with differences in the arrangements of male and female gonophores in the two clades. In Clade A, male and female zooids are found within the same cormidium (a single reiterated sequence of zooids along the stem, see Fig. 2). In these species, the male and female zooids are placed at different but well defined locations within the cormidium. Meanwhile in calycophorans, each cormidium bears either male or female gonophores. In this form of monoecy, the male and female cormidia can either occur in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. In sum, homoplasy in sexual-system evolution along with variation in the spatial arrangement of gonophores within a colony suggest that siphonophores have evolved different ways to be monoecious. The sexual system and cormidial arrangement of *Rudjakovia* is undescribed, although preliminary observations suggest that this species may be monoecious and that monoecy arose a third time in the Pyrostephidae. A detailed redescription of *Rudjakovia* would help clarify this.

@@

### 3.4 The Evolution of Zooid Types

@@

Palpons are modified reduced gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. We do not distinguish here between gonopalpons (palpons associated with gonodendra, without a tentacle, as in the cystonects) and palpons borne on the stem (typically with a reduced tentacle or palpacle) [@totton1965synopsis]. We reconstruct them as present in the common ancestor of siphonophores (Fig. 5B), retained in most species, but lost three times independently in the branches leading to Pyrostephidae (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans, and in *Marrus claudanielis*. Within the calycophorans, one species *Stephanophyes superba* (not included in this phylogeny) has polyp-like zooids that have been described as palpons [@totton1965synopsis], but the exact identity of this zooid is not clear and needs further morphological examination.

@@

### 3.5 The Gain and Loss of the Pneumatophore

The pneumatophore is a gas-filled float located at the anterior end of the colony, which helps the colony maintain its orientation in the water column, and plays a role in flotation in the case of the cystonects [@church2015histology;@mackie1974locomotion; @totton1965synopsis]. It is not a zooid, as it is not formed by budding but by invagination at the aboral end of the planula during early development [@garstang1946morphology; @leloup1935siphonophores; @carre1969etude]. Recent descriptions of the neural arrangement in the pneumatophore of *Nanomia bijuga*  suggest it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis [@church2015histology]. The ancestral siphonophore had a pneumatophore (Fig. 2B), since both cystonects and all "physonects" possess one (Fig. 4). The pneumatophore was lost in Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy [@mackie1974locomotion].

### 3.6 The Gain and Loss of Tentilla

@@

### 3.7 The Evolution of Vertical Habitat Use

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). Sequencing at the Brown Genomics Core facility was supported in part by NIH P30RR031153 and NSF EPSCoR EPS-1004057. Data transfer was supported by NSF RII-C2 EPS-1005789. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. SOWHAT analyses were carried out on the Odyssey cluster supported by the FAS Division of Science, Research Computing Group at Harvard University -- we thank Cassandra Extavour for use of the Harvard cluster. We thank Rob Sherlock for providing the *Bathyphysa conifera* photograph. We also thank Zack Lewis for help with RNA extractions. We also thank the MBARI crews and ROV pilots for collection of the specimens.

>>>>

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at most depths of the ocean - from the surface, like the familiar Portuguese man of war, to the deep sea. Siphonophores play important roles in ocean ecosystems, and are among the most abundant gelatinous predators. A previous phylogenetic study based on two ribosomal RNA genes provided insight into the internal relationships between major siphonophore groups, however there was little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny based on new transcriptome data from 29 siphonophore species analyzed in combination with 14 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including sexual system (monoecy vs. dioecy), gain and loss of zooid types, life history traits, and habitat. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but we find strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose at least twice.

@@

Siphonophores (Fig. 1 and 2) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical; @purcell1981dietary; @pages2001gelatinous;  @Choy20172116]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at most depths in the ocean, with the deepest recorded species found around 4300m [@lindsay2005planktonic]. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be more than 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, Rhodaliidae, that are tethered to the bottom for part of their lives [@pugh1983benthic]. There are currently 187 valid described siphonophore species [@worms].

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years due to their unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones (Fig. 2) then arise and asexually produce other genetically identical zooids that remain attached [@carre1969etude; @carre1991complete; @carre1967; @carre1993ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from other colonial animals in their colony structure and development -- their zooids are highly functionally specialized and arranged in precise, repeating, species-specific patterns [@Beklemishev1969;@cartwright2010character]. Zooids are specialized for a range of functions, including feeding, reproduction, or swimming (Fig. 2) [@dunn2006evolution].

@@

The phylogenetic relationships recovered in this study received strong support across analysis methods (Fig. 3A), with a couple of localized exceptions (Fig. 3B). Most clades are consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. We provisionally refer here to Pyrostephidae as the clade including *Rudjakovia* sp., although sampling of *Pyrostephos vanhoeffeni* is needed in order to determine if *Rudjakovia* sp. falls within Pyrostephidae or is sister to it. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

@@

```{r char_evo_setup, include=F, echo=FALSE}
#Load data
read.csv('character_coding/main_characters.csv', header = T, sep = ',') -> cdata
rownames(cdata) = cdata$Species

@@

### 3.2 Evolution of Monoecy

@@

Ancestral character state reconstructions favor the hypothesis that monoecy arose twice (Fig. 5A), once in Calycophorae and once in Clade A. This is consistent with differences in the arrangements of male and female gonophores in the two clades. In Clade A, male and female zooids are found within the same cormidium (a single reiterated sequence of zooids along the stem, see Fig. 2). In these species, the male and female zooids are placed at different but well defined locations within the cormidium. Meanwhile in calycophorans, each cormidium bears either male or female gonophores. In this form of monoecy, the male and female cormidia can either occur in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. One exception to this is Abylids, where both male or female gonophores may be found within the same eudoxid [@carre1967]. In sum, homoplasy in sexual-system evolution along with variation in the spatial arrangement of gonophores within a colony suggest that siphonophores have evolved different ways to be monoecious. The sexual system and cormidial arrangement of *Rudjakovia* is undescribed, although preliminary observations suggest that this species may be monoecious and that monoecy arose a third time in the Pyrostephidae. A detailed redescription of *Rudjakovia* would help clarify this.

@@

### 3.3 The Evolution of Zooid Types

@@

Palpons are modified reduced gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. We do not distinguish here between gonopalpons (palpons associated with gonodendra, without a tentacle, as in the cystonects) and palpons borne on the stem (typically with a reduced tentacle or palpacle) [@totton1965synopsis]. We reconstruct them as present in the common ancestor of siphonophores (Fig. 5B), retained in most species, but lost three times independently in the branches leading to Pyrostephidae (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans, and in *Marrus claudanielis* and Undescribed sp. L. Within the calycophorans, one species *Stephanophyes superba* (not included in this phylogeny) has polyp-like zooids that have been described as palpons [@totton1965synopsis], but the exact identity of this zooid is not clear and needs further morphological examination.

@@

### 3.4 The Gain and Loss of the Pneumatophore

The pneumatophore (Fig. 2A) is a gas-filled float located at the anterior end of the colony, which helps the colony maintain its orientation in the water column, and plays a role in flotation in the case of the cystonects [@church2015histology;@mackie1974locomotion; @totton1965synopsis]. It is not a zooid, as it is not formed by budding but by invagination at the aboral end of the planula during early development [@garstang1946morphology; @leloup1935siphonophores; @carre1969etude]. Recent descriptions of the neural arrangement in the pneumatophore of *Nanomia bijuga*  suggest it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis [@church2015histology]. The ancestral siphonophore had a pneumatophore (Fig. 2B), since both cystonects and all "physonects" possess one (Fig. 4). The pneumatophore was lost in Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy [@mackie1974locomotion].

### 3.5 The Gain and Loss of Tentilla

@@

### 3.6 The Evolution of Vertical Habitat Use

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). Sequencing at the Brown Genomics Core facility was supported in part by NIH P30RR031153 and NSF EPSCoR EPS-1004057. Data transfer was supported by NSF RII-C2 EPS-1005789. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. SOWHAT analyses were carried out on the Odyssey cluster supported by the FAS Division of Science, Research Computing Group at Harvard University -- we thank Cassandra Extavour for use of the Harvard cluster. We thank Rob Sherlock for providing the *Bathyphysa conifera* photograph. We also thank Zack Lewis for help with RNA extractions and the MBARI crews and ROV pilots for collection of the specimens.
